### 2024-06-20 manuscript.Rmd
some minor fixes added thanks to Leonie

authornote: |
  Please address correspondence to: Prof. Dr. Andreas M. Brandmaier, Rüdesheimerstr. 50, 14197 Berlin, Germany. Email: andreas.brandmaier@medicalschool-berlin.de.
  The R package described in this article can be downloaded from https://github.com/brandmaier/reproducibleRchunks. The manuscript source code is available at https://github.com/brandmaier/reproducibleRchunks-manuscript.

@@

The `reproducibleRchunk` package allows users to make computational results in R Markdown documents automatically testable for reproduction (does the same script with the same data produce the same results, e.g. on a different computer and/or later in time) with minimal changes to your workflow.
There is only a single thing users need to change in their Markdown document if they are already using R Markdown: In the first code chunk of their document, load the package and then change the code chunk type from `r` to `reproducibleR` for all code chunks that should be automatically checked for reproducibility.
In `reproducibleR` chunks, all newly declared variables are automatically identified and their contents are stored in a meta data file. 
Furthermore, the code chunk itself (that is, the code syntax) is also fingerprinted and stored, such that later non-reproduction can be traced back to either changes of the code chunk or changes of the computation.
Once a document is reproduced, in R-Markdown-speak _knitted_ again [@yihui2015], all computational results generated in the `reproducibleR` chunks are automatically tested for reproducibility.
The result of the reproduction attempt is displayed in a reproducibility report for each chunk.
Customizations of these reports are possible as we will demonstrate later.
A simple way to assess reproducibility of computational results would be  a bitwise comparison of the entire knitted document (i.e., the resulting HTML, PDF or DOCX file).
Yet, a particular advantage of using `reproducibleR` chunks is that users have a fine-grained control over which stages of a data analysis or any other computation should be checked for reproducibility. 
For example, if there is non-reproducibility, is it due to changes in the data file, during data cleaning or statistical modeling. 
Using `reproducibleRchunks`, users can separately assess the reproducibility of each stage of a given statistical data analysis (or any other computation).
Also, not all elements of a research asset are required to reproduce (such as the current date displayed in a presentation of results).
In the following, we will briefly demonstrate how successful and failed reproduction will be checked and displayed by the package, we will go into details of the mechanics of the package, and discuss potential cases of non-reproducibility and how the package will support researchers in handling these cases.

@@

In the following, we will describe how the package stores, retrieves and compares reproducibility information.
First of all, the package executes `reproducibleR` code chunks just as regular R code chunks.
That is, there are no additional limits imposed on what can be computed and tested for reproducibility.
Once code execution is completed, the package gathers information about all variables that were newly declared in the current chunk. 
The contents of those variables are stored in a separate JSON data file.
JSON is a shorthand for Java Script Object Notation and its main features are that it is an open standard for mapping complex objects to text files with high simplicity and readability for machines and humans [@lennon2009introduction].
The name of the JSON file contains the original Markdown file and the chunk label and always starts with the prefix `repro`). 
That is, if reproducibility information of a chunk labelled `datacleaning` in the file `sem_analysis.Rmd` is stored in a file called _.repro_sem_analysis.Rmd_datacleaning.json_.
Once the document is re-generated and JSON data files exist, their content is checked against the newly computed chunk variables for identity.

```{r echo=FALSE, eval=TRUE}
set.seed(42)
numbers_ex <- sample(1:10, 5)
```
Here is an example of how the contents of a single variable called `numbers` is stored in JSON format. Here, the variable content is a vector of five random draws from the number one to ten: `r numbers_ex` generated by the following command:

```{r echo=TRUE, eval=FALSE}
numbers_ex <- sample(1:10, 5)
```

@@

\begin{verbatim}
```{r echo=FALSE, eval=TRUE, results="asis"}
content <- jsonlite::serializeJSON(numbers_ex, pretty=TRUE)
content <- gsub(pattern="\\n",replacement="  \n", x=content)
cat(content)
```
\end{verbatim}

For privacy reasons and to save disk space, we actually do not store the raw data by default but only fingerprints of the data.
Fingerprints are realized via so-called one-way hash functions, which can take an arbitrary large digital object and map it onto a fixed-size object, which is typically displayed as hexadecimal string.
Fingerprints allow for an efficient comparison of identity of results while avoid data security and privacy issues as they do not allow to reproduce the original data.
The following shows how the information about variable `numbers` is stored using a SHA256-fingerprint (which is a 256 bit-long fingerprint that is usually displayed using 64 hexadecimal characters):

\begin{verbatim}
```{r echo=FALSE, eval=TRUE, results="asis"}
content <- jsonlite::serializeJSON(digest::digest(numbers_ex,algo = "sha256"),pretty=TRUE)
content <- gsub(pattern="\\n",replacement="  \n", x=content)
cat(content)
```
\end{verbatim}

@@

Once a document is (re-)generated with reproducibility data present (i.e., reproducibleRchunks with matching JSON files), the package will compare all objects amenable for reproducibility checks.
Identity of original and reproduced results are checked with `identical()` function from R's `base` package, which according to the R documentation is a "safe and reliable way to test two objects (even complex ones) for being exactly identical" (see `?identical` R documentation).
Using exact comparisons of object identity is sometimes more restrictive than one would expect.
This is particularly true for numeric comparisons. 
For example results of numeric computations may be logically equivalent but not bitwise equivalent; in numeric representations, there is sometimes a bit reserved for the sign of a number, thus it can be possible to have -0 and +0, which both are exactly zero in a decimal representation but the binary representations differ.
Numeric representations may also differ up to a given numeric precision, for example, when comparing reproducibility across machines that work with different numerical precision (32bit vs 64bit). To avoid such problems, `reproducibleRchunks` rounds numeric values to a given precision (by default, eight digits)

>>>>

authornote: |
  Please address correspondence to: Prof. Dr. Andreas M. Brandmaier, Rüdesheimerstr. 50, 14197 Berlin, Germany. Email: andreas.brandmaier@medicalschool-berlin.de.
  The R package described in this article can be downloaded from https://github.com/brandmaier/reproducibleRchunks. The manuscript source code is available at https://github.com/brandmaier/reproducibleRchunks-manuscript. We thank Leonie Hagitte for providing comments on an earlier version of the manuscript.

@@

The `reproducibleRchunk` package allows users to make computational results in R Markdown documents automatically testable for reproduction (does the same script with the same data produce the same results, e.g. on a different computer and/or later in time) with minimal changes to your workflow.
There is only a single thing users need to change in their Markdown document if they are already using R Markdown: In the first code chunk of their document, load the package and then change the code chunk type from `r` to `reproducibleR` for all code chunks that should be automatically checked for reproducibility.
In `reproducibleR` chunks, all newly declared variables are automatically identified and their contents are stored in a meta data file. 
Furthermore, the code chunk itself (that is, the code syntax) is also fingerprinted and stored, such that later non-reproduction can be traced back to either changes of the code chunk or changes of the computation.
Once a document is reproduced, in R-Markdown-speak _knitted_ again [@yihui2015], all computational results generated in the `reproducibleR` chunks are automatically tested for reproducibility.
The result of the reproduction attempt is displayed in a reproducibility report for each chunk.
Customizations of these reports are possible as we will demonstrate later.
A simple way to assess reproducibility of computational results would be  a bitwise comparison of the entire knitted Markdown document (i.e., the resulting HTML, PDF or DOCX file) to a previous version.
Yet, a particular advantage of using `reproducibleR` chunks is that users have a fine-grained control over which stages of a data analysis or any other computation should be checked for reproducibility. 
For example, if there is non-reproducibility, is it due to changes in the data file, during data cleaning or statistical modeling. 
Using `reproducibleRchunks`, users can separately assess the reproducibility of each stage of a given statistical data analysis (or any other computation).
Also, not all elements of a research asset may be required to reproduce (such as the current date displayed in a presentation of results).
In the following, we will briefly demonstrate how successful and failed reproduction will be checked and displayed by the package, we will go into details of the mechanics of the package, and discuss potential cases of non-reproducibility and how the package will support researchers in handling these cases.

@@

In the following, we will describe how the package stores, retrieves and compares reproducibility information.
First of all, the package executes `reproducibleR` code chunks just as regular R code chunks.
That is, there are no additional limits imposed on what can be computed and tested for reproducibility.
Once code execution is completed, the package gathers information about all variables that were newly declared in the current chunk. 
The contents of those variables are stored in a separate JSON data file.
JSON is a shorthand for Java Script Object Notation and its main features are that it is an open standard for mapping complex objects to text files with high simplicity and readability for machines and humans [@lennon2009introduction].
The name of the JSON file contains the original Markdown file and the chunk label and, by package default, starts with the prefix `.repro`. 
That is, if reproducibility information of a chunk labelled `datacleaning` in the file `sem_analysis.Rmd` is stored in a file called _.repro_sem_analysis.Rmd_datacleaning.json_.
Once the document is re-generated and JSON data files exist, their content is checked against the newly computed chunk variables for identity.

```{r echo=FALSE, eval=TRUE}
set.seed(42)
numbers <- sample(1:10, 5)
```
Here is an example of how the contents of a single variable called `numbers` is stored in JSON format. Here, the variable content is a vector of five random draws from the number one to ten: `r numbers` generated by the following command:

```{r echo=TRUE, eval=FALSE}
set.seed(42)
numbers <- sample(1:10, 5)
```

@@

\begin{verbatim}
```{r echo=FALSE, eval=TRUE, results="asis"}
content <- jsonlite::serializeJSON(numbers, pretty=TRUE)
content <- gsub(pattern="\\n",replacement="  \n", x=content)
cat(content)
```
\end{verbatim}

The JSON format has a clear, formal structure, which allows easy parsing of the information both for computers and humans. 
Without going into detail, we believe it suffices to say that for each object, its type (character, integer, list, class, etc.), its value and its attributes are stored whereas values itself can again be collections of objects; in this way, arbitrarily complex R objects (such as an entire regression model) can be represented.
For privacy reasons and to save disk space, we actually do not store the raw data by default but only fingerprints of the data.
Fingerprints are realized via so-called one-way hash functions, which can take an arbitrary large digital object and map it onto a fixed-size object, which is typically displayed as hexadecimal string.
Fingerprints allow for an efficient comparison of identity of results while avoid data security and privacy issues as they do not allow to reproduce the original data.
The following shows how the information about variable `numbers` is stored using a SHA256-fingerprint (which is a 256 bit-long fingerprint that is usually displayed using 64 hexadecimal characters):

\begin{verbatim}
```{r echo=FALSE, eval=TRUE, results="asis"}
content <- jsonlite::serializeJSON(digest::digest(numbers,algo = "sha256"),pretty=TRUE)
content <- gsub(pattern="\\n",replacement="  \n", x=content)
cat(content)
```
\end{verbatim}

@@

Once a document is (re-)generated with reproducibility data present (i.e., reproducibleRchunks with matching JSON files), the package will compare all objects amenable for reproducibility checks.
Identity of original and reproduced results are checked with `identical()` function from R's `base` package, which according to the R documentation is a "safe and reliable way to test two objects (even complex ones) for being exactly identical" (see `?identical` R documentation).
Using exact comparisons for testing object identity is sometimes more restrictive than one would expect.
This is particularly true for numeric comparisons. 
For example results of numeric computations may be logically equivalent but not bitwise equivalent; in numeric representations, there is sometimes a bit reserved for the sign of a number, thus it can be possible to have -0 and +0, which both are exactly zero in a decimal representation but the binary representations differ.
Numeric representations may also differ up to a given numeric precision, for example, when comparing reproducibility across machines that work with different numerical precision (32bit vs 64bit). To avoid such problems, `reproducibleRchunks` rounds numeric values to a given precision (by default, eight digits)
